---
title: "SOCAT-GLODAP"
author: "Jean-Pierre Gattuso & Robert Schlegel"
date: '`r format(Sys.Date(), "%d %B %Y")`'
# site: workflowr::wflow_site
# output:
#   workflowr::wflow_html:
#     toc: false
output:
  rmarkdown::html_document:
    theme: paper
    number_sections: false
# editor_options:
#   chunk_output_type: console
# csl: frontiers.csl
# bibliography: FACE-IT.bib
---

```{r setup, message = FALSE, warning = FALSE, echo = FALSE}
# Start with common project code
Sys.setlocale("LC_ALL", "en_US.UTF-8")
Sys.setenv(TZ = 'UTC') # on utilise UTC
rm(list = ls())

# Remove scientific notation
options(scipen = 9999)

# Find who is the user and define the pCloud path
if (Sys.getenv("LOGNAME") == "gattuso") {
  pCloud_path = "~/pCloud\ Drive/"
} else {
  pCloud_path = "~/pCloudDrive/" 
}

library(tidyverse)
library(data.table) # For 'fread'
library(doParallel) # For using multiple cores
library(tidync)

# Load global marine borders according to International Hydrographic Organization (1953)
# See https://doi.pangaea.de/10.1594/PANGAEA.777975?format=html#download
marine_borders <- read_delim("metadata/Limits_of_oceans_and_seas.tab", delim = "\t", skip = 16, 
                             col_names = c("name", "lat", "lon", "index_name", "index_point"))
marine_borders %>% filter(lon >= 80, lon<=150, lat >= -30, lat<= 30) %>% 
    ggplot()+
  geom_polygon(aes(x = lon, y = lat, fill = name)) +
  borders(fill = "grey30")

  
# Get only the borders for the oceans
# NB: Keeping the eastern and western partitions of oceans to avoid issues with longitude wrapping
# NB: There is no Southern Ocean
ocean_borders <- marine_borders %>% 
  filter(grepl("ocean|Mediterranean|Baltic", name, ignore.case = T)) %>% 
  filter(!grepl("north|south|region", name, ignore.case = T)) %>% 
#  filter(!grepl("south", name, ignore.case = T)) %>% 
  mutate(lon = case_when(lon >= 179.98 ~ 180.0,
                         lon <= -179.98 ~ -180.0,
                         TRUE ~ lon))

# Make a few specific tweaks to tighten up the polygons
# This pulls up the corners of the Arctic polygons so that they capture the full ~60 - 90 degree range
# ocean_borders$lat[305] <- 90
# ocean_borders$lat[268] <- 90
# ocean_borders$lon[266] <- 0
# ocean_borders$lon[269] <- 0
# ocean_borders$lon[307] <- 0

# Test visuals
ggplot(data = ocean_borders) +
  geom_polygon(aes(x = lon, y = lat, fill = name)) +
  borders(fill = "grey30")

# Function for finding and cleaning up points within a given region polygon
points_in_region <- function(region_name, data_product, border_product){
  region_sub <- border_product %>% 
    filter(name == region_name)
  coords_in <- data_product %>% 
    mutate(in_grid = sp::point.in.polygon(point.x = data_product[["lon"]], point.y = data_product[["lat"]], 
                                          pol.x = region_sub[["lon"]], pol.y = region_sub[["lat"]])) %>% 
    filter(in_grid >= 1) %>% 
    mutate(name = region_name) %>% 
    dplyr::select(lon, lat, name)
  return(coords_in)
}
```

# Overview

The world ocean is 362,033,000 km2 (Menard & Smith, 1966 in Jakobsson et al., 2002). According to Jakobsson et al. (2002) the Arctic Ocean, as defined by the International Hydrographic Organization (1953), is 15,551,000 km2 or about 4.3% of the global ocean. 

In the present analysis, the Arctic Ocean is as defined by the International Hydrographic Organization (1953) using the following dataset:

International Hydrographic Organization, IHO; Sieger, Rainer (2012): Limits of oceans and seas in digitized, machine readable form. Alfred Wegener Institute, Helmholtz Centre for Polar and Marine Research, Bremerhaven, PANGAEA, https://doi.org/10.1594/PANGAEA.777975

Calculation of the surface area of the Arctic Ocean using GMT (script below) was finally not used:

- north of 60°N: 1.6751e+07 km2, or `r round(100  * 1.6751e+07 / 361132000, digits = 1)`% of the global ocean.
- north of 75°N: 7.11213e+06 km2, or `r round(100  * 7.11213e+06 / 361132000, digits = 1)`% of the global ocean.

# SOCAT Product

```{r read-socat, message = FALSE, warning = FALSE, echo = FALSE}
# Read from pCloudDrive (veryyy slow!)
# socat <- data.table::fread(file = paste0(pCloud_path, "FACE-IT_data/socat/SOCATv2021.tsv"), 
#                            select = c("yr", "mon", "day", "longitude [dec.deg.E]", "latitude [dec.deg.N]", 
#                                       "sample_depth [m]", "ETOPO2_depth [m]", "pCO2water_SST_wet [uatm]"), 
#                            skip = 6626, nrows = Inf, nThread = 10, fill = TRUE)
socat <- readRDS(file = paste0(pCloud_path, "FACE-IT_data/socat/SOCATv2021.rds")) %>% # whole ocean
  dplyr::filter(!is.na(`pCO2water_SST_wet [uatm]`)) %>% # remove NAs
  dplyr::rename(year = yr, month = mon, lon = `longitude [dec.deg.E]`, lat = `latitude [dec.deg.N]`, 
                bottomdepth = `ETOPO2_depth [m]`, depth = `sample_depth [m]`, pco2 = `pCO2water_SST_wet [uatm]`) %>% 
  mutate(product = "SOCAT",
         depth = replace_na(depth, 0),
         lon = case_when(lon > 180 ~ lon -360, TRUE ~ lon)) %>% # Correct longitude range to match ocean border polygons
  filter(depth >= 0)

# Find marine borders for the SOCAT product and merge
registerDoParallel(cores = 15)
socat_regions <- plyr::ldply(unique(ocean_borders$name), points_in_region, .parallel = T, 
                             data_product = distinct(socat[,c("lon", "lat")]), border_product = ocean_borders)

# Process data
socat_all <- socat %>% # whole ocean
  dplyr::left_join(socat_regions, by = c("lon", "lat")) %>%
  dplyr::filter(!is.na(pco2)) %>%  # remove NAs
  dplyr::mutate(depth = replace_na(depth, 0)) %>% 
  dplyr::filter(depth >= 0) %>%
  dplyr::rename(ocean = name) %>%
  dplyr::mutate(
    product = "SOCAT",
    ocean = gsub(", western part", "", ocean),
    ocean = gsub(", eastern part", "", ocean),
    ocean = case_when(
      lat <= -60 ~ "Southern Ocean",
      is.na(ocean) ~ "Not Ocean",
      TRUE ~ ocean)) %>% 
  dplyr::mutate(ocean = case_when(ocean == "Not Ocean" & lon <= 160 & lon >= 90 & lat >= -30 & lat <= 30~ "East Indian Archipelago", TRUE ~ ocean))
socat_all$ocean <- as.factor(socat_all$ocean)
summary(socat_all$ocean) #problem: there are 9596678 "Not ocean" data (more than 60% of the data)! 
# The issue was that SOCAT has longitude from 0 to 360, but the ocean boundaries are -180 to 180

# test plot
socat_all %>% 
  mutate(lon = round(lon), lat = round(lat)) %>% 
  group_by(lon, lat, ocean) %>% 
  summarise(count = n(), .groups = "drop") %>% 
  ggplot(aes(x = lon, y = lat, colour = ocean)) +
  borders(fill = "grey30") +
  geom_point() +
  coord_cartesian(expand = F)

# Make Arctic cuts
socat_arctic <- socat_all %>% 
  dplyr::filter(ocean == "Arctic Ocean")
socat_arctic_coastal <- socat_all %>% 
  dplyr::filter(ocean == "Arctic Ocean",
                bottomdepth < 200)

# timeline
dat_yr <- socat_arctic %>% count(year)
timel_socat_arctic_plt <- dat_yr %>% 
  ggplot(aes(x = year, y = n)) +
  geom_point() +
  geom_line() +
  labs(title = "socat: number of Arctic data per year and <= 200m", x = "Year", y = "tco2")
print(timel_socat_arctic_plt)

# monthly distribution Arctic
dat_mon_arctic <- socat_arctic %>% count(month) %>% 
  right_join(data.frame(month = 1:12), by = "month") %>% 
  mutate(n = replace_na(n, 0))
month_socat_arctic_plt <- dat_mon_arctic %>% 
  ggplot() +
  geom_col(aes(x = month, y = n, fill = month), show.legend = F) +
  scale_x_continuous(breaks = 1:12, labels = month.abb, expand = c(0, 0)) +
  labs(title = "socat: number of Arctic data per month", x = "Month", y = "pCO2water_SST_wet")
print(month_socat_arctic_plt)
socat_arctic_summer <- dat_mon_arctic %>% 
  dplyr::filter(month >= 6 & month <= 9) %>%
  sum()
socat_arctic_summer_pc <- round(100 * socat_arctic_summer / sum(dat_mon_arctic$n), digits = 1)

# monthly distribution coastal (< 200 m) Arctic
dat_mon_arctic_coastal <- socat_arctic_coastal %>% count(month) %>% 
  right_join(data.frame(month = 1:12), by = "month") %>% 
  mutate(n = replace_na(n, 0))
month_socat_arctic_coastal_plt <- dat_mon_arctic_coastal %>% 
  ggplot() +
  geom_col(aes(x = month, y = n, fill = month), show.legend = F) +
  scale_x_continuous(breaks = 1:12, labels = month.abb, expand = c(0, 0)) +
  labs(title = "socat: number of coastal (< 200 m) Arctic data per month", x = "Month", y = "pCO2water_SST_wet")
print(month_socat_arctic_coastal_plt)
socat_arctic_coastal_summer <- dat_mon_arctic_coastal %>% 
  dplyr::filter(month >= 6 & month <= 9) %>%
  sum()
socat_arctic_coastal_summer_pc <- round(100 * socat_arctic_coastal_summer / sum(dat_mon_arctic_coastal$n), digits = 1)
```

SOCAT v2021 has `r nrow(socat_all)` records of pCO2water_SST_wet [uatm], including:

* `r nrow(socat_arctic)` in the Arctic Ocean, or `r round(100*nrow(socat_arctic)/nrow(socat_all), digits = 1)`% of the global number of records
* `r nrow(socat_arctic_coastal)` in the Arctic Ocean shelf waters (shallower than 200 m), or `r round(100*nrow(socat_arctic_coastal)/nrow(socat_arctic), digits = 1)`% of the number of records in the Arctic.
* `r socat_arctic_summer_pc`% of all Arctic data have been collected over 4 months (June to September).
* `r socat_arctic_coastal_summer_pc`% of the shelf (< 200m) data have been collected over 4 months (June to September).


# GLODAP product

```{r read-glodap, message=FALSE, warning=FALSE, echo = FALSE}
# The bottle data
# This is slow
# GLODAP_bottle <- data.table::fread(paste0(pCloud_path,"FACE-IT_data/glodap/GLODAPv2.2021_Merged_Master_File.csv"), na.strings = "-9999") 
GLODAP_bottle <- tibble(readRDS(file = paste0(pCloud_path, "FACE-IT_data/socat/GLODAP_bottle.rds"))) %>%
  rename_with(~gsub("G2", "", .x, fixed = TRUE)) %>% 
  dplyr::select(year, month, day, longitude, latitude, bottomdepth, depth, tco2) %>% 
  dplyr::rename(lon = longitude, lat = latitude) 

# Find marine borders for GLODAP product
registerDoParallel(cores = 15)
glodap_regions <- plyr::ldply(unique(ocean_borders$name), points_in_region, .parallel = T, 
                             data_product = distinct(GLODAP_bottle[,c("lon", "lat")]), border_product = ocean_borders)

# Process data
glodap_all <- GLODAP_bottle %>% # whole ocean
  dplyr::left_join(glodap_regions, by = c("lon", "lat")) %>%
  dplyr::filter(!is.na(tco2)) %>%  # remove NAs
  dplyr::filter(depth >= 0) %>%
  dplyr::rename(ocean = name) %>%
  dplyr::mutate(
    product = "GLODAP",
    ocean = gsub(", western part", "", ocean),
    ocean = gsub(", eastern part", "", ocean),
    ocean = case_when(
      lat <= -60 ~ "Southern Ocean",
      is.na(ocean) ~ "Not Ocean",
      TRUE ~ ocean)) %>%
  dplyr::mutate(ocean = case_when(ocean == "Not Ocean" & lat <= 48.5 & lat >= 29 ~ "Mediterranean", TRUE ~ ocean))
glodap_all$ocean <- as.factor(glodap_all$ocean)
summary(glodap_all$ocean) #there are 1374 "Not ocean" data (0.2% of all data)
# These are data points in seas, mostly the Mediterranean

# test plot
glodap_all %>% 
  mutate(lon = round(lon), lat = round(lat)) %>% 
  group_by(lon, lat, ocean) %>% 
  summarise(count = n(), .groups = "drop") %>% 
  ggplot(aes(x = lon, y = lat, colour = ocean)) +
  borders(fill = "grey30") +
  geom_point() +
  coord_cartesian(expand = F)

# Make Arctic cut
glodap_arctic <- glodap_all %>% 
  dplyr::filter(ocean == "Arctic Ocean")
glodap_arctic_coastal <- glodap_all %>% 
  dplyr::filter(ocean == "Arctic Ocean",
                bottomdepth < 200)

# timeline
dat_yr <- glodap_arctic %>% count(year)
timel_glodap_arctic_plt <- dat_yr %>% 
  ggplot(aes(x = year, y = n)) +
  geom_point() +
  geom_line() +
  labs(title = "GLODAP: number of Arctic data per year", x = "Year", y = "tco2")
print(timel_glodap_arctic_plt)


# monthly distribution Arctic
dat_mon_arctic <- glodap_arctic %>% count(month) %>% 
  right_join(data.frame(month = 1:12), by = "month") %>% 
  mutate(n = replace_na(n, 0))
month_glodap_arctic_plt <- dat_mon_arctic %>% 
  ggplot() +
  geom_col(aes(x = month, y = n, fill = month), show.legend = F) +
  scale_x_continuous(breaks = 1:12, labels = month.abb, expand = c(0, 0)) +
  labs(title = "glodap: number of Arctic data per month", x = "Month", y = "pCO2water_SST_wet")
print(month_glodap_arctic_plt)
glodap_arctic_summer <- dat_mon_arctic %>% 
  dplyr::filter(month >= 6 & month <= 9) %>%
  sum()
glodap_arctic_summer_pc <- round(100 * glodap_arctic_summer / sum(dat_mon_arctic$n), digits = 1)

# monthly distribution coastal (< 200 m) Arctic
dat_mon_arctic_coastal <- glodap_arctic_coastal %>% count(month) %>% 
  right_join(data.frame(month = 1:12), by = "month") %>% 
  mutate(n = replace_na(n, 0))
month_glodap_arctic_coastal_plt <- dat_mon_arctic_coastal %>% 
  ggplot() +
  geom_col(aes(x = month, y = n, fill = month), show.legend = F) +
  scale_x_continuous(breaks = 1:12, labels = month.abb, expand = c(0, 0)) +
  labs(title = "glodap: number of coastal (< 200 m) Arctic data per month", x = "Month", y = "pCO2water_SST_wet")
print(month_glodap_arctic_coastal_plt)
glodap_arctic_coastal_summer <- dat_mon_arctic_coastal %>% 
  dplyr::filter(month >= 6 & month <= 9) %>%
  sum()
glodap_arctic_coastal_summer_pc <- round(100 * glodap_arctic_coastal_summer / sum(dat_mon_arctic_coastal$n), digits = 1)
```

GLODAP v2.2021 has `r nrow(glodap_all)` records of TCO2, including:

* `r nrow(glodap_arctic)` in the Arctic Ocean, or `r round(100*nrow(glodap_arctic)/nrow(glodap_all), digits = 1)`% of the total number of records
* `r nrow(glodap_arctic_coastal)` in the Arctic Ocean shelf waters (shallower than 200 m) or `r round(100*nrow(glodap_arctic_coastal)/nrow(glodap_arctic), digits = 1)`% of the total number of records in the Arctic
* `r glodap_arctic_summer_pc`% of all Arctic data have been collected over 4 months (June to September).
* `r glodap_arctic_coastal_summer_pc`% of the shelf (< 200m) data have been collected over 4 months (June to September).


# Comparison

With the Arctic region investigated for both products we now want to compare them for the Arctic, Atlantic, Indian, Pacific, and Southern oceans. For both the full oceans as well as coastal (<= 200 m bottom depth) values. To do so we will use the bottom depths accompanied by the products.

```{r comparison, message=FALSE, warning=FALSE, echo = FALSE}
# Combine products for ease of plotting
both_all <- bind_rows(glodap_all, socat_all)

# Rough visual of the pixels by ocean
both_all %>% 
  # These lines reduce the count of points for plotting
  mutate(lon = round(lon), lat = round(lat)) %>% 
  group_by(lon, lat, ocean, product) %>% 
  summarise(n = n(), .groups = "drop") %>% 
  # NB: Without the above lines R may hang
  ggplot(aes(x = lon, y = lat)) +
  borders() +
  geom_point(aes(colour = ocean)) +
  coord_cartesian(expand = F) +
  labs(x = NULL, y = NULL) +
  facet_wrap(~product)

# Calculate annual counts
annual_count_deep <- both_all %>% 
  group_by(product, ocean, year) %>% 
  summarise(n = n(), .groups = "drop") %>% 
  mutate(depth_group = "all")
annual_count_shallow <- both_all %>% 
  filter(bottomdepth <= 200) %>% 
  group_by(product, ocean, year) %>% 
  summarise(n = n(), .groups = "drop") %>% 
  mutate(depth_group = "shallow")
annual_count <- rbind(annual_count_deep, annual_count_shallow)

# Dot and line plot of count of data per year per ocean etc.
annual_count %>% 
  ggplot(aes(x = year, y = n, colour = ocean)) +
  geom_point(aes(shape = product)) +
  geom_line(aes(linetype = product)) +
  labs(title = "Number of data per year for different oceans", 
       subtitle = "Note there is so much more SOCAT data it is hard to see GLODAP", 
       x = "Year", y = "count") +
  facet_wrap(~depth_group) +
  theme(legend.position = "bottom")

# Calculate monthly counts
monthly_count_deep <- both_all %>% 
  group_by(product, ocean, month) %>% 
  summarise(n = n(), .groups = "drop") %>% 
  mutate(depth_group = "all")
monthly_count_shallow <- both_all %>% 
  filter(bottomdepth <= 200) %>% 
  group_by(product, ocean, month) %>% 
  summarise(n = n(), .groups = "drop") %>% 
  mutate(depth_group = "shallow")
monthly_count <- rbind(monthly_count_deep, monthly_count_shallow)

# Bar plot showing monthly count comparisons
monthly_count %>% 
  ggplot() +
  geom_col(aes(x = month, y = n, fill = ocean, colour = product)) +
  scale_colour_brewer(palette = "Set1") +
  scale_x_continuous(breaks = 1:12, labels = month.abb, expand = c(0, 0)) +
  labs(title = "Count of data per month for different oceans",
       subtitle = "Note that there is so little GLODAP data by comparison it is difficult to see",
       x = "Month", y = "count") +
  facet_wrap(~depth_group) +
  theme(legend.position = "bottom")
```

# Appendix: GMT script

Works but results not used in the manuscript in the end.

```
——
bash
cd
cd "pCloud Drive/FACE-IT_data/maps/GEBCO/"

# masque terre
gmt grdlandmask -Gnewmask.nc -RGEBCO_2020.nc -Df -N1/NaN/NaN/NaN/NaN

# selection de l ocean a partir du masques
gmt grdmath GEBCO_2020.nc newmask.nc OR = wet0.nc
# il reste des valeurs positives que j'élimine
gmt grdclip wet0.nc -Gwet.nc -Sa0/NaN

# calcul de la surface de l ocean, latitude > 75 degrés
#  la valeur -10000 que j'ai donnée à l'option -C de grdvolume
#  est une profondeur (en valeur absolue supérieure à la profondeur maximale
#  de l'océan), ce qui permet, si j'ai bien compris le fonctionnement
#  de grdvolume) de prendre en compte toute la surface de l'océan
gmt grdvolume wet.nc -R-180/180/75/90 -C-10000 -Sk > grdv_wet_75

# calcul de la surface de l ocean, latitude > 60 degrés
gmt grdvolume wet.nc -R-180/180/60/90 -C-10000 -Sk > grdv_wet_60.txt

# la surface est la somme des surfaces des 2 demi-hémisphères
#   (la surface est la 2ième valeur sortie par grdvolume)
#   (l unité de surface est le km2)
cat grdv_wet_75_dh? | awk 'BEGIN{s=0};{s=s+$2};END{print s}'
# 7.11213e+06
# Moi je trouve 7112242
cat grdv_wet_60_dh? | awk 'BEGIN{s=0};{s=s+$2};END{print s}'
# 1.6751e+07
# Moi je trouve 16768436

# j ai calculé pour latitude > 60 afin de comparer avec la surface
#  de l océan arctique donnée par wikipédia : 1.4e+7 km2
```