---
title: "SOCAT-GLODAP"
author: "Jean-Pierre Gattuso"
date: '`r format(Sys.Date(), "%d %B %Y")`'
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: false
# output:
#   rmarkdown::html_document:
#     theme: paper
#     number_sections: false
editor_options:
  chunk_output_type: console
csl: frontiers.csl
bibliography: FACE-IT.bib
---

```{r setup, message = FALSE, warning = FALSE, echo = FALSE}
# Start with common project code
Sys.setlocale("LC_ALL", "en_US.UTF-8")
Sys.setenv(TZ = 'UTC') # on utilise UTC
rm(list = ls())

# Remove scientific notation
options(scipen = 9999)

# Find who is the user and define the pCloud path
if (Sys.getenv("LOGNAME") == "gattuso") {
  pCloud_path = "~/pCloud\ Drive/"
} else {
  pCloud_path = "~/pCloudDrive/" 
}
library(tidyverse)
library(data.table) # For 'fread'
library(doParallel) # For using multiple cores
library(tidync)

# Load global marine borders
marine_borders <- read_delim("metadata/Limits_of_oceans_and_seas.tab", delim = "\t", skip = 16, 
                             col_names = c("name", "lat", "lon", "index_name", "index_point"))

# Get only the borders for the oceans
# NB: Keeping the eastern and western partitions of oceans to avoid issues with longitude wrapping
# NB: There is no Southern Ocean
ocean_borders <- marine_borders %>% 
  filter(grepl("ocean", name, ignore.case = T)) %>% 
  filter(!grepl("north", name, ignore.case = T)) %>% 
  filter(!grepl("south", name, ignore.case = T))

# Test visuals
ggplot(data = ocean_borders) +
  borders() +
  geom_polygon(aes(x = lon, y = lat, fill = name))

# Function for finding and cleaning up points within a given region polygon
points_in_region <- function(region_name, data_product, border_product){
  region_sub <- border_product %>% 
    filter(name == region_name)
  coords_in <- data_product %>% 
    mutate(in_grid = sp::point.in.polygon(point.x = data_product[["lon"]], point.y = data_product[["lat"]], 
                                          pol.x = region_sub[["lon"]], pol.y = region_sub[["lat"]])) %>% 
    filter(in_grid >= 1) %>% 
    mutate(name = region_name) %>% 
    dplyr::select(lon, lat, name)
  return(coords_in)
}
```

# Overview

The world ocean is 361,132,000 km2 (https://en.wikipedia.org/wiki/Ocean#Physical_properties). According to Wikipedia the Arctic Ocean is about 14,056,000 km2. Calculation of the surface area of the Arctic Ocean using GMT (script below):

- north of 60°N: 1.6751e+07 km2, or `r round(100  * 1.6751e+07 / 361132000, digits = 1)`% of the global ocean.
- north of 75°N: 7.11213e+06 km2, or `r round(100  * 7.11213e+06 / 361132000, digits = 1)`% of the global ocean.

# SOCAT Product

```{r read socat, message = FALSE, warning = FALSE, echo = FALSE}
# The full SOCAT product
# NB: This takes a few minutes due to the size of the file
socat <- data.table::fread(file = paste0(pCloud_path, "FACE-IT_data/socat/SOCATv2021.tsv"), 
                           select = c("yr", "mon", "day", "longitude [dec.deg.E]", "latitude [dec.deg.N]", 
                                      "sample_depth [m]", "ETOPO2_depth [m]", "pCO2water_SST_wet [uatm]"), 
                           skip = 6626, nrows = Inf, nThread = 10, fill = TRUE) %>% 
  dplyr::rename(year = yr, month = mon, lon = `longitude [dec.deg.E]`, lat = `latitude [dec.deg.N]`, 
                  bottomdepth = `ETOPO2_depth [m]`, depth = `sample_depth [m]`, tco2 = `pCO2water_SST_wet [uatm]`) %>% 
  mutate(lon = case_when(lon > 180 ~ lon-360, TRUE ~ lon))

# Find marine borders for the SOCAT product
registerDoParallel(cores = 15)
socat_regions <- plyr::ldply(unique(ocean_borders$name), points_in_region, .parallel = T, 
                             data_product = distinct(socat[,c("lon", "lat")]), border_product = ocean_borders)

# Clean up product for analyses
socat_all <- socat %>% # whole ocean
  dplyr::filter(!is.na(tco2)) %>% # remove NAs
  mutate(product = "SOCAT",
         depth = replace_na(depth, 0)) %>% 
  filter(depth >= 0) %>% 
  left_join(socat_regions, by = c("lon", "lat")) %>% 
  mutate(name = case_when(lat <= -60 ~ "Southern Ocean", TRUE ~ name))

# Male some simple cuts to data
socat_arctic_60 <- socat_all %>% 
  dplyr::filter(lat > 60, bottomdepth <= 200)
socat_arctic_75 <- socat_all %>% 
  dplyr::filter(lat > 75, bottomdepth <= 200)

# Timeline
socat_dat_yr <- socat_arctic_75 %>% count(year)
timel_socat_a75_plt <- socat_dat_yr %>% 
  ggplot(aes(x = year, y = n)) +
  geom_point() +
  geom_line() +
  labs(title = "SOCAT: number of data per year at > 75°N and <= 200m", x = "Year", y = "pCO2water_SST_wet")
print(timel_socat_a75_plt)

# monthly distribution
socat_dat_mon <- socat_arctic_75 %>% count(month) %>% 
  right_join(data.frame(month = 1:12), by = "month") %>% 
  mutate(n = replace_na(n, 0))
month_socat_a75_plt <- socat_dat_mon %>% 
  ggplot() +
  geom_col(aes(x = month, y = n, fill = month), show.legend = F) +
  scale_x_continuous(breaks = 1:12, labels = month.abb, expand = c(0, 0)) +
  labs(title = "SOCAT: number of data per month at > 75°N and <= 200m", x = "Month", y = "pCO2water_SST_wet")
print(month_socat_a75_plt)
socat_summer <- socat_dat_mon %>% 
  dplyr::filter(month >= 6 & month <=9) %>%
  sum()
socat_summer_pc <- round(100 * socat_summer / sum(socat_dat_mon$n), digits = 1)
```

SOCAT v2021 has `r nrow(socat_all)` records of pCO2water_SST_wet [uatm], including:

* `r nrow(socat_arctic_60)` in the Arctic Ocean shelf waters (North of 60°N and shallower than 200 m), or `r round(100*nrow(socat_arctic_60)/nrow(socat_all), digits = 1)`% of the total number of records
* `r nrow(socat_arctic_75)` in the Arctic Ocean shelf waters (North of 75°N and shallower than 200 m), or `r round(100*nrow(socat_arctic_75)/nrow(socat_all), digits = 1)`% of the total number of records
* `r socat_summer_pc`% of the data have been collected over 4 months (June to September) and almost no data were collected from November (`r socat_dat_mon$n[socat_dat_mon$month == 11]`) to April (0).

# GLODAP product

```{r read-glodap, message=FALSE, warning=FALSE, echo = FALSE}
# The bottle data
glodap_bottle <- data.table::fread(paste0(pCloud_path,"FACE-IT_data/glodap/GLODAPv2.2021_Merged_Master_File.csv"), na.strings = "-9999", nThread = 10) %>%
  rename_with(~gsub("G2", "", .x, fixed = TRUE)) %>% 
  dplyr::select(year, month, day, longitude, latitude, bottomdepth, depth, tco2) %>% 
  dplyr::rename(lon = longitude, lat = latitude)

# Find marine borders for GLODAP product
registerDoParallel(cores = 15)
glodap_regions <- plyr::ldply(unique(ocean_borders$name), points_in_region, .parallel = T, 
                             data_product = distinct(glodap_bottle[,c("lon", "lat")]), border_product = ocean_borders)

# Clean up the GLODAP product
glodap_all <- glodap_bottle %>% # whole ocean
  dplyr::filter(!is.na(tco2)) %>%  # remove NAs
  mutate(product = "GLODAP") %>% 
  filter(depth >= 0) %>% 
  left_join(glodap_regions, by = c("lon", "lat")) %>% 
  mutate(name = case_when(lat <= -60 ~ "Southern Ocean", TRUE ~ name))

# Make some quick cuts
glodap_arctic_60 <- glodap_bottle %>% 
  dplyr::filter(lat > 60, bottomdepth <= 200)
glodap_arctic_75 <- glodap_bottle %>% 
  dplyr::filter(lat > 75, bottomdepth <= 200)

# timeline
glodap_dat_yr <- glodap_arctic_75 %>% count(year)
timel_glodap_a75_plt <- glodap_dat_yr %>% 
  ggplot(aes(x = year, y = n)) +
  geom_point() +
  geom_line() +
  labs(title = "GLODAP: number of data per year at > 75°N and <= 200m", x = "Year", y = "tco2")
print(timel_glodap_a75_plt)

# monthly distribution
glodap_dat_mon <- glodap_arctic_75 %>% count(month) %>% 
  right_join(data.frame(month = 1:12), by = "month") %>% 
  mutate(n = replace_na(n, 0))
month_glodap_a75_plt <- glodap_dat_mon %>% 
  ggplot() +
  geom_col(aes(x = month, y = n, fill = month), show.legend = F) +
  scale_x_continuous(breaks = 1:12, labels = month.abb, expand = c(0, 0)) +
  labs(title = "GLODAP: number of data per month at > 75°N", x = "Month", y = "pCO2water_SST_wet")
print(month_glodap_a75_plt)
glodap_summer <- glodap_dat_mon %>% 
  dplyr::filter(month >= 6 & month <= 9) %>%
  sum()
glodap_summer_pc <- round(100 * glodap_summer / sum(glodap_dat_mon$n), digits = 1)
```

GLODAP v2.2021 has `r nrow(glodap_all)` records of TCO2, including:

* `r nrow(glodap_arctic_60)` in the Arctic Ocean shelf waters (North of 60°N and shallower than 200 m), or `r round(100*nrow(glodap_arctic_60)/nrow(glodap_all), digits = 1)`% of the total number of records
* `r nrow(glodap_arctic_75)` in the Arctic Ocean shelf waters (North of 75°N  and shallower than 200 m) or `r round(100*nrow(glodap_arctic_75)/nrow(glodap_all), digits = 1)`% of the total number of records
* `r glodap_summer_pc`% of the data have been collected over 4 months (June to September) and no data have been collected from November to March.

# Comparison

With the Arctic region investigated for both products we now want to compare them for the Arctic, Atlantic, Indian, Pacific, and Southern oceans. For both the full oceans as well as coastal (<= 200 m bottom depth) values. To do so we will use the bottom depths accompanied by the products, and will define very rough bounding boxes for the oceans.

```{r comparison, message=FALSE, warning=FALSE, echo = FALSE}
# Combine products for ease of plotting
both_all <- rbind(glodap_all, socat_all)

# Rough visual of the pixels by ocean
both_all %>% 
  # These lines reduce the count of points for plotting
  mutate(lon = round(lon), lat = round(lat)) %>% 
  group_by(lon, lat, name, product) %>% 
  summarise(n = n(), .groups = "drop") %>% 
  # NB: Without the above lines R may hang
  ggplot(aes(x = lon, y = lat)) +
  borders() +
  geom_point(aes(colour = name)) +
  coord_cartesian(expand = F) +
  labs(x = NULL, y = NULL) +
  facet_wrap(~product)
# NB: Note the large missing portions of the Arctic Ocean

# Calculate annual counts
annual_count_deep <- both_all %>% 
  group_by(product, name, year) %>% 
  summarise(n = n(), .groups = "drop") %>% 
  mutate(depth_group = "all")
annual_count_shallow <- both_all %>% 
  filter(bottomdepth <= 200) %>% 
  group_by(product, name, year) %>% 
  summarise(n = n(), .groups = "drop") %>% 
  mutate(depth_group = "shallow")
annual_count <- rbind(annual_count_deep, annual_count_shallow)

# Dot and line plot of count of data per year per ocean etc.
annual_count %>% 
  ggplot(aes(x = year, y = n, colour = name)) +
  geom_point(aes(shape = product)) +
  geom_line(aes(linetype = product)) +
  labs(title = "Number of data per year for different oceans", 
       subtitle = "Note there is so much more SOCAT data it is hard to see GLODAP", 
       x = "Year", y = "count") +
  facet_wrap(~depth_group) +
  theme(legend.position = "bottom")

# Calculate monthly counts
monthly_count_deep <- both_all %>% 
  group_by(product, name, month) %>% 
  summarise(n = n(), .groups = "drop") %>% 
  mutate(depth_group = "all")
monthly_count_shallow <- both_all %>% 
  filter(bottomdepth <= 200) %>% 
  group_by(product, name, month) %>% 
  summarise(n = n(), .groups = "drop") %>% 
  mutate(depth_group = "shallow")
monthly_count <- rbind(monthly_count_deep, monthly_count_shallow)

# Bar plot showing monthly count comparisons
monthly_count %>% 
  ggplot() +
  geom_col(aes(x = month, y = n, fill = name, colour = product)) +
  scale_colour_brewer(palette = "Set1") +
  scale_x_continuous(breaks = 1:12, labels = month.abb, expand = c(0, 0)) +
  labs(title = "Count of data per month for different oceans",
       subtitle = "Note that there is so little GLODAP data by comparison it is difficult to see",
       x = "Month", y = "count") +
  facet_wrap(~depth_group) +
  theme(legend.position = "bottom")
```

