---
title: "SOCAT-GLODAP"
author: "Jean-Pierre Gattuso"
date: '`r format(Sys.Date(), "%d %B %Y")`'
output:
  rmarkdown::html_document:
    theme: paper
    number_sections: false
---

```{r setup, message = FALSE, warning = FALSE, echo = FALSE}
# Start with common project code
# Find who is the user and define the pCloud path
Sys.setlocale("LC_ALL", "en_US.UTF-8")
Sys.setenv(TZ='UTC') # on utilise UTC
rm(list = ls())
if (Sys.getenv("LOGNAME") == "gattuso"){
  pCloud_path = "~/pCloud\ Drive/"
} else {
  pCloud_path = "~/pCloudDrive/" 
}
library(tidyverse)
library(data.table) # For 'fread'
library(doParallel) # For using multiple cores
library(tidync)
library(workflowr)
```
The world ocean is 361,132,000 km2 (https://en.wikipedia.org/wiki/Ocean#Physical_properties). According to Wikipedia the Arctic Ocean is about 14,056,000 km2. Calculation of the surface area of the Arctic Ocean using GMT (script below):

- north of 60°N: 1.6751e+07 km2, or `r round(100  * 1.6751e+07 / 361132000, digits = 1)`% of the global ocean.
- north of 75°N: 7.11213e+06 km2, or `r round(100  * 7.11213e+06 / 361132000, digits = 1)`% of the global ocean.


```{r read socat, message = FALSE, warning = FALSE, echo = FALSE}
socat <- data.table::fread(file = paste0(pCloud_path, "FACE-IT_data/socat/SOCATv2020.tsv"), select = c("yr", "mon", "day", "longitude [dec.deg.E]", "latitude [dec.deg.N]", "pCO2water_SST_wet [uatm]"), skip = 6362, nrows = Inf, nThread = 10, fill = TRUE)
socat_all <- socat %>% # whole ocean
  dplyr::filter(!is.na(`pCO2water_SST_wet [uatm]`)) # remove NAs
socat_arctic_60 <- socat_all %>% 
  dplyr::filter(`latitude [dec.deg.N]` > 60, !is.na(`pCO2water_SST_wet [uatm]`))
socat_arctic_75 <- socat_all %>% 
  dplyr::filter(`latitude [dec.deg.N]` > 75, !is.na(`pCO2water_SST_wet [uatm]`))

# timeline
dat <- socat_arctic_75 %>% count(yr)
timel_socat_a75_plt <- dat %>% 
  ggplot() +
  geom_point(aes(x = yr, y = n)) +
  labs(title = "SOCAT: number of data per year at > 75°N", x = "Year", y = "pCO2water_SST_wet")
print(timel_socat_a75_plt)

# monthly distribution
dat <- socat_arctic_75 %>% count(mon)
month_socat_a75_plt <- dat %>% 
  ggplot() +
  geom_col(aes(x = mon, y = n)) +
  labs(title = "SOCAT: number of data per month at > 75°N", x = "Year", y = "pCO2water_SST_wet")
print(month_socat_a75_plt)
socat_summer <- dat %>% 
  dplyr::filter(mon >= 6 & mon <=9) %>%
  sum()
socat_summer_pc <- round(100 * socat_summer / sum(dat$n), digits = 1)
```

SOCAT v2020 has `r nrow(socat_all)` records of pCO2water_SST_wet [uatm], including:

* `r nrow(socat_arctic_60)` in the Arctic Ocean (North of 60°N), or `r round(100*nrow(socat_arctic_60)/nrow(socat_all), digits = 1)`% of the total number of records
* `r nrow(socat_arctic_75)` in the Arctic Ocean (North of 75°N), or `r round(100*nrow(socat_arctic_75)/nrow(socat_all), digits = 1)`% of the total number of records

`r socat_summer_pc`% of the data have been collected over 4 months (June to September) and no data was collected in December and January.

```{r read-glodap, , message=FALSE, warning=FALSE, echo = FALSE}

# The bottle data
GLODAP_bottle <- read_csv(paste0(pCloud_path,"FACE-IT_data/glodap/GLODAPv2.2020_Merged_Master_File.csv")) %>%
  dplyr::select(year, month, day, longitude, latitude, bottomdepth, depth, tco2, talk)

glodap_all <- GLODAP_bottle %>% # whole ocean
  dplyr::filter(!is.na(tco2)) # remove NAs
glodap_arctic_60 <- GLODAP_bottle %>% 
  dplyr::filter(latitude > 60, !is.na(tco2))
glodap_arctic_75 <- GLODAP_bottle %>% 
  dplyr::filter(latitude > 75, !is.na(tco2))

# timeline
dat <- glodap_arctic_75 %>% count(year)
timel_glodap_a75_plt <- dat %>% 
  ggplot() +
  geom_point(aes(x = year, y = n)) +
  labs(title = "GLODAP: number of data per year at > 75°N", x = "Year", y = "tco2")
print(timel_glodap_a75_plt)

# monthly distribution
dat <- glodap_arctic_75 %>% count(month)
month_glodap_a75_plt <- dat %>% 
  ggplot() +
  geom_col(aes(x = month, y = n)) +
  labs(title = "GLODAP: number of data per month at > 75°N", x = "Year", y = "pCO2water_SST_wet")
print(month_glodap_a75_plt)
glodap_summer <- dat %>% 
  dplyr::filter(month >= 6 & month <=9) %>%
  sum()
glodap_summer_pc <- round(100 * glodap_summer / sum(dat$n), digits = 1)
```

GLODAP v2.2016 has `r nrow(glodap_all)` records of TCO2, including:

* `r nrow(glodap_arctic_60)` in the Arctic Ocean (North of 60°N), or `r round(100*nrow(glodap_arctic_60)/nrow(glodap_all), digits = 1)`% of the total number of records
* `r nrow(glodap_arctic_75)` in the Arctic Ocean (North of 75°N) or `r round(100*nrow(glodap_arctic_75)/nrow(glodap_all), digits = 1)`% of the total number of records

`r glodap_summer_pc`% of the data have been collected over 4 months (June to September) and no data was collected in December and January.
